---
title: Ant data Generalized Linear Model - Bayesian
author: Brett Melbourne
date: 5 Nov 2024
output: github_document
---

Third in a series of scripts to analyze the ant data described in Ellison (2004). This script includes specifying a Bayesian GLM with Poisson distribution and log link, training by HMC, and calculating uncertainties for derived quantities to answer our three scientific questions. Here we code the model explicitly with the function `ulam()` from McElreath's `rethinking` package and code posterior samples for derived quantities directly from the samples of parameters. Future scripts will consider likelihood and frequentist GLMs with the `lme4` package, and easy Bayesian fitting of GLMs with the `rstanarm` package, and multilevel models to fully account for the design structure.

This Rmarkdown script can be rendered to a reproducible report (Ctrl+Shift+K in RStudio).

```{r results=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(rethinking)
theme_set(theme_bw())
```

Read in and plot the data

```{r}
ant <- read.csv("data/ants.csv")
ant$habitat <- factor(ant$habitat)
ant |> 
    ggplot(mapping=aes(x=latitude, y=richness, col=habitat)) +
    geom_point()
```

## Model

For the linear predictor, we'll write the model out explicitly, with the data organized into single-column variables that express a linear-additive model design matrix, as in the class slides.

$$
\begin{align}
  y_i &\sim \mathrm{Poisson}(\mu_i) && \\
  log(\mu) &= \beta_0 \mathrm{intercept}_i + 
              \beta_1 \mathrm{forest}_i + 
              \beta_2 \mathrm{latitude}_i + 
              \beta_3 \mathrm{forest}_i \mathrm{latitude}_i && \\
  \beta_0 &\sim \mathrm{Normal}(1, 10) && \\
  \beta_1 &\sim \mathrm{Normal}(0, 10) && \\
  \beta_2 &\sim \mathrm{Normal}(0, 10) && \\
  \beta_3 &\sim \mathrm{Normal}(0, 10) && \\
\end{align}
$$
I've chosen (very) weakly informative priors, similar to those we've seen in McElreath so far. For $\beta_0$, I've centered the prior at 1. The linear predictor is on the log scale, so 1 corresponds to $e^1 = 2.7$ for bog species richness, which is closer to the mean richness of the data than $e^0 = 1$ and is more sensibly shaped than a prior with a lot of probability mass near a species richness of 0.

## Training

Bayesian fit with `ulam`, spelling out the model. We first need new columns to represent the design matrix. We'll also scale by latitude (center and divide by sd) to improve efficiency of the training algorithm.

```{r}
# Set up variables
mean_lat <- mean(ant$latitude)
sd_lat <- sd(ant$latitude)
d <- ant |>
    mutate(intercept = rep(1, n()),
           forest = ifelse(habitat == "forest", 1, 0),
           latitude = (latitude - mean_lat) / sd_lat,
           forest_X_latitude = forest * latitude) |>
    select(richness, intercept, forest, latitude, forest_X_latitude)
d
```

Now specify the model in full and train with `ulam`. It will take a minute to compile, then it will sample from the posterior. I'm choosing to compile and run with one chain at first to check that the model compiles, then if all goes well, we'll carry on.

```{r cache=TRUE}
bysfitHxL <- ulam(
    alist(
        richness ~ dpois(mu),
        log(mu) <- beta_0 * intercept +
                   beta_1 * forest +
                   beta_2 * latitude +
                   beta_3 * forest_X_latitude,
        beta_0 ~ dnorm(1, 10),
        beta_1 ~ dnorm(0, 10),
        beta_2 ~ dnorm(0, 10),
        beta_3 ~ dnorm(0, 10)
    ),
    data=d
)
```

The model compiles and the training algorithm doesn't complain. So, now for a serious run. There will be a bunch of messages tracking progress of the chains.

```{r message=FALSE, cache=TRUE}
bysfitHxL <- ulam(bysfitHxL, chains=4, cores=4, warmup=1000, iter=10000)
```

Parameter information

```{r}
precis(bysfitHxL, prob=0.95, digits=4)
```

In the `precis` table, we can look at the posterior sample means (`mean`) for a point estimate of the parameter values. The standard deviation (`sd`) of the posterior samples is one measure of uncertainty. The range between the 2.5 and 97.5 percentiles is a 95% central posterior interval (CPI) for the parameters. The final two columns are MCMC diagnostics. We see especially that the `rhat` values are all 1, indicating convergence, and `ess_bulk` (effective number of samples) is very high for all parameters. Remember that these parameters describe the linear predictor, which is on the log scale. For example, $\beta_0$ is the intercept for bog habitat and backtransformed corresponds to about 4.5 species, which is in the right ballpark for the data with latitude centered.


### Working with posterior samples

We'll first work with the samples directly (as we did in McElreath Ch 4). Doing it the slightly harder way first will prepare you for building your own custom analyses by developing a deeper understanding of the data structures as well as experience with visualization. In the next script, we'll do the same analysis with convenient functions in `rstanarm`.

```{r}
samples <- extract.samples(bysfitHxL)
class(samples)
str(samples)
names(samples)
```


### Visualizing the posterior

We can plot histograms directly from the samples. To do this with `ggplot`, we first need the samples in a dataframe. We can go one step further, converting the dataframe to tidy format (using `pivot_longer` from `tidyr`) and then using `facet_wrap` to plot histograms for all 4 parameters. First, make a copy of the samples in a dataframe data structure (instead of the default list structure).

```{r}
samplesdf <- data.frame(samples)
head(samplesdf)
```

Plot the histograms

```{r}
samplesdf |> 
    pivot_longer(cols=everything(), names_to="parameter", values_to="sample_value") |> 
    ggplot() +
    geom_histogram(mapping=aes(x=sample_value, y=after_stat(density)), bins=75) +
    facet_wrap(vars(parameter), scales="free")
```

These histograms are not merely diagnostic. *The posterior distribution is the full inference for a parameter*. Diagnostically, we want to ask if the samples are giving us a good picture of the posterior distribution. These look very nice. We see that the distributions are roughly symmetric (statistical theory for these types of models suggests the posterior would be approximately Normal) and quite smooth (not noisy) due to the large number of samples.

Sometimes you might only want one of the histograms, in which case you can do it quickly and succinctly by using `hist` from base plot (not shown).

```{r eval=FALSE}
hist(samples$beta_0, breaks=75)
```


Here's a graphical summary of the parameters using the `rethinking` plot method. This shows 89% intervals by default.

```{r}
plot(bysfitHxL)
```

Check the traceplot to confirm convergence and check the behavior.

```{r}
traceplot(bysfitHxL)
```

We'll dive into more diagnostic checks for MCMC chains and the model in a later script.


## Inference

### Parameter credible intervals

We get 95% credible intervals (HPDI) directly from the samples

```{r}
HPDI(samples$beta_0, prob=0.95)
HPDI(samples$beta_1, prob=0.95)
HPDI(samples$beta_2, prob=0.95)
HPDI(samples$beta_3, prob=0.95)
```

These are almost the same as the CPIs (in the `precis` table above) due to the symmetric posteriors and our large set of samples.


### Mean curves, intervals for curves (HPDI), posterior predictive distribution

These quantities don't come directly from the samples for individual parameters but instead are quantities that we derive from combinations of the samples (i.e. functions of the samples). The following is quite literal. We could make this more elegant but the steps needed are clear this way.

```{r}
# Initialize a grid of latitudes, scaled the same as we scaled the data
lat_upr <- (45 - mean_lat) / sd_lat
lat_lwr <- (41.92 - mean_lat) / sd_lat
latitude <- seq(from=lat_lwr, to=lat_upr, length.out=50)

# Initialize storage
n <- length(latitude)
hpdi_bog <- matrix(NA, nrow=n, ncol=5) #to store hpdi values and mean
colnames(hpdi_bog) <- c("mnmu","mulo95","muhi95","ppdlo95","ppdhi95")
hpdi_forest <- matrix(NA, nrow=n, ncol=5)
colnames(hpdi_forest) <- c("mnmu","mulo95","muhi95","ppdlo95","ppdhi95")

# For each latitude, form the posterior
for ( i in 1:n ) {
    
    # First form samples for the linear predictor \eta
    eta_bog <- samples$beta_0 + 
               samples$beta_2 * latitude[i]
    eta_forest <- samples$beta_0 + 
                  samples$beta_1 + 
                  samples$beta_2 * latitude[i] + 
                  samples$beta_3 * latitude[i]
    
    # Then use inverse link for samples of the posterior \mu
    mu_bog <- exp(eta_bog)
    mu_forest <- exp(eta_forest)
    
    # Sample from Poisson to get the posterior predictive distribution
    ppd_bog <- rpois(n=length(mu_bog), lambda=mu_bog)
    ppd_forest <- rpois(n=length(mu_forest), lambda=mu_forest)
    
    # Mean and intervals of these samples
    hpdi_bog[i,1] <- mean(mu_bog)
    hpdi_bog[i,2:3] <- HPDI(mu_bog, prob=0.95)
    #hpdi_bog[i,4:5] <- HPDI(ppd_bog, prob=0.95)
    hpdi_bog[i,4:5] <- quantile(ppd_bog, prob=c(0.025,0.975)) #CPI
    hpdi_forest[i,1] <- mean(mu_forest)
    hpdi_forest[i,2:3] <- HPDI(mu_forest, prob=0.95)
    #hpdi_forest[i,4:5] <- HPDI(ppd_forest, prob=0.95)
    hpdi_forest[i,4:5] <- quantile(ppd_forest, prob=c(0.025,0.975)) #CPI
    
}
rm(eta_bog, eta_forest, mu_bog, mu_forest) #clean up
```

Notice that we calculated expectations (e.g. `mean(mu_bog)`) and intervals directly on the scale of the data (the "response" scale), not on the linear predictor scale. If we calculated first on the linear predictor scale and then backtransformed the HPDI intervals to the response scale they would be biased due to nonlinear averaging. Also, the posterior predictive distribution (PPD) can, of course, only be on the response scale. I used the CPI (`quantile()`) for the posterior predictive distribution because plots of the HPDI and CPI were substantially similar but the CPI was more numerically stable.

Package in tidy format for plotting

```{r}
latitude <- latitude * sd_lat + mean_lat
predsbog <- data.frame(habitat=rep("bog", n), latitude, hpdi_bog)
predsforest <- data.frame(habitat=rep("forest", n), latitude, hpdi_forest)
preds <- rbind(predsbog, predsforest)
rm(latitude, n, hpdi_bog, hpdi_forest, predsbog, predsforest) #clean up
```

Now we can visualize these inferences. The plot is built in layers with different `geoms` representing different aspects of the model (curves, credible intervals, prediction intervals) and data. We'll add two variations to our previous visualizations of these data to improve visual perception and interpretation (i.e. to make the data and story clear):

1. colors with semantic meaning (green for forest, orange/brown for bog) that
are also colorblind friendly
2. labels added directly to the plot to reduce translational effort

The credible intervals for the means are the shaded regions while the dashed lines show the posterior predictive interval.

```{r}
bfc <- c("#d95f02", "#1b9e77") #bog & forest colors
preds |>
    ggplot() +
    geom_ribbon(mapping=aes(x=latitude, ymin=mulo95, ymax=muhi95, fill=habitat),
                alpha=0.2) +
    geom_point(data=ant, mapping=aes(x=latitude, y=richness, col=habitat)) +
    geom_line(mapping=aes(x=latitude, y=mnmu, col=habitat)) +
    geom_line(mapping=aes(x=latitude, y=ppdlo95, col=habitat), lty=2) +
    geom_line(mapping=aes(x=latitude, y=ppdhi95, col=habitat), lty=2) +
    annotate("text", x=42.7, y=3.3, label="Bog", col=bfc[1]) +
    annotate("text", x=43.85, y=9.5, label="Forest", col=bfc[2]) +
    scale_fill_manual(values=bfc) +
    scale_color_manual(values=bfc) +
    scale_y_continuous(breaks=seq(0, 20, 4), minor_breaks=seq(0, 20, 2)) +
    xlab("Latitude (degrees north)") +
    ylab("Ant species richness") +
    theme(legend.position="none")
```

Notice that the intervals for forest are wider than for bog. This is because the uncertainty scales with the mean of the response. Also notice that the intervals for the posterior predictive distribution have discrete steps. This is because the data generating process is discrete (e.g. we cannot have 1.3 species). The intervals are quite smooth. With fewer iterations, Monte Carlo error would be greater and we would see a few blips in the intervals for the predictive distribution and some wiggles in the mean intervals.

Let's consider now our original three scientific questions. We can answer question 1: **How does species richness vary with latitude?** The above plot shows this. Species richness declines with latitude and appears to be well described by a gentle exponential decline over the range of the data (because the inverse link function is exponential). The uncertainty in this relationship for both habitats is shown by the credible intervals for the mean. We can also answer question 2: **Is this relationship different between habitats?** There are two answers, depending on our view. On the one hand, we could conclude from the plot that species richness starts higher and declines faster with latitude in forest than in bog habitat. On the other hand, the exponential decline is about the same and there is little evidence for a difference in the exponent as judged by the value and uncertainty of $\beta_3$ (the difference in the exponent between habitats). The mean and 95% CPI for this parameter is (repeating here from earlier):

```{r}
mean(samples$beta_3)
HPDI(samples$beta_3, prob=0.95)
```

Finally, we can't properly answer question 3: **How different is species richness between habitats?** We can see from the plot above that mean species richness is mostly higher in forest than bog, judging by the non-overlap of the credible intervals for the mean. We can also roughly read off the difference (about 5.5 at latitude 42, or 2.5 at latitude 45). But we haven't yet precisely quantified this difference or its uncertainty. Let's do that next.

### Non-standard derived quantities

To answer question 3, we need to make our own relevant quantities: the differences in mean species richness between habitats at different latitudes. These differences are a function of the parameters, so we can **derive** samples of them from samples of the parameters. The first half of this code is substantially the same as above; the main action is at the line that calculates the difference. The object `diff` contains the posterior samples for the difference in mean richness at a given altitude. I found the CPI to be a good and stable estimate of the credible interval.

```{r}
# Initialize variables and storage 
lat_lwr <- (41.92 - mean_lat) / sd_lat
lat_upr <- (45 - mean_lat) / sd_lat
latitude <- seq(from=lat_lwr, to=lat_upr, length.out=50)
n <- length(latitude)
forest_bog_diff <- matrix(NA, nrow=n, ncol=3) #to store mean and hpdi values
colnames(forest_bog_diff) <- c("mndiff","difflo95","diffhi95")

# For each latitude, form the posterior
for ( i in 1:n ) {
    
    # First form samples for the linear predictor \eta
    eta_bog <- samples$beta_0 + 
               samples$beta_2 * latitude[i]
    eta_forest <- samples$beta_0 + 
                  samples$beta_1 + 
                  samples$beta_2 * latitude[i] + 
                  samples$beta_3 * latitude[i]
    
    # Then use inverse link for samples of the posterior \mu
    mu_bog <- exp(eta_bog)
    mu_forest <- exp(eta_forest)
    
    # Now calculate the habitat difference (derived quantity)
    diff <- mu_forest - mu_bog
    
    # Mean and intervals of these samples
    forest_bog_diff[i,1] <- mean(diff)
    forest_bog_diff[i,2:3] <- HPDI(diff, prob=0.95)
    #forest_bog_diff[i,2:3] <- quantile(diff, prob=c(0.025,0.975)) #CPI

}

# Package in a dataframe (with latitude rescaled to original)
latitude <- latitude * sd_lat + mean_lat
diff_df <- data.frame(cbind(forest_bog_diff, latitude))
rm(latitude,n,forest_bog_diff,eta_bog,eta_forest,mu_bog,mu_forest,diff) #clean up
```

Plot the difference with its uncertainty. I've used `coord_cartesian()` to set the y-axis limits instead of `ylim()` in case the ribbon region goes beyond the axis limit. In `ggplot()` the dataset is truncated to the axis limits by default, which could lead to a missing chunk of the ribbon. The upper limit of the interval is very close to 8.3 and goes a hair above in some runs of the stochastic sampler.

```{r}
diff_df |> 
    ggplot() +
    geom_ribbon(mapping=aes(x=latitude, ymin=difflo95, ymax=diffhi95),
        alpha=0.2) +
    geom_line(mapping=aes(x=latitude, y=mndiff)) +
    coord_cartesian(ylim=c(0,8.3)) +
    xlab("Latitude (degrees north)") +
    ylab("Difference in species richness (forest - bog)")
```

Now we have quantitatively answered question 3: **How different is species richness between habitats?** We can see how the difference declines with latitude, and we can see how the uncertainty changes with latitude showing that mean ant richness in forest is clearly higher than in bog across the entire range of latitudes.