---
title: Ant data Generalized Linear Model - frequentist
author: Brett Melbourne
date: 11 Oct 2018 (updated 24 Oct 2022)
output:
    github_document
---

Fourth in a series of scripts to analyze the ant data described in Ellison (2004). This script demonstrates frequentist inference from the Poisson-log link GLM (model training is by maximum likelihood). Previous scripts have considered visualization, fitting linear-Normal models with likelihood and SSQ, model checking, and Bayesian inference. Here we fit the model using the function `glm()` in the `stats` package of base R, which uses model syntax like `lm()`. The next script will again fit the model using a Bayesian approach but with the analogous `stan_glm()` function from the `rstanarm` package. Future scripts will consider multilevel models to fully account for the design structure and data variance.

```{r results=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
theme_set(theme_bw())
```

Read in and plot the data

```{r}
ant <- read.csv("data/ants.csv")
ant$habitat <- factor(ant$habitat)
ant |> 
    ggplot(mapping=aes(x=latitude, y=richness, col=habitat)) +
    geom_point()
```

## Model

The model is the same as in the previous Bayesian script, except now we don't have priors for the parameters.

$$
\begin{align}
  y_i &\sim \mathrm{Poisson}(\mu_i) && \\
  log(\mu) &= \beta_0 \mathrm{intercept}_i + 
              \beta_1 \mathrm{forest}_i + 
              \beta_2 \mathrm{latitude}_i + 
              \beta_3 \mathrm{forest}_i \mathrm{latitude}_i &&
\end{align}
$$

## Training

Train the model using `glm()`. The training algorithm is a specialized algorithm for maximum likelihood estimation of exponential distribution-family models called iterative weighted least squares (Nelder & Wedderburn 1972), a variant of the Newton-Raphson method.

```{r}
fitHxL <- glm(richness ~ habitat + latitude + habitat:latitude, family=poisson, data=ant)
coef(fitHxL) #Maximum likelihood estimates of parameters
```

Model checking

```{r}
par(mfrow=c(2,2))
plot(fitHxL,1:4, ask=FALSE)
```

Overall, the diagnostics seem not too bad. It is certainly an improvement over the Normal linear model we tried before. There is perhaps still a slight tendency for the standardized residuals to increase with the mean (we'll investigate that further later) and there is a still the same influential point (case 25). We should investigate the influential point using our case deletion algorithm to examine parameter sensitivity to this point. We'll return to that later as we continue to improve the model.


## Inference

Out of the box inferences about parameter uncertainty come from the assumption of a Normal sampling distribution for the parameter z-values (estimate divided by standard error), giving us a p-value for each parameter.

```{r}
summary(fitHxL)
```
```{r eval=FALSE}
coef(summary(fitHxL))[,1:2]   #or use this for just the coefficients
```

However, most of the parameter hypotheses being considered here are not meaningful, either because we don't care about them biologically (e.g. we don't care about the `intercept` hypothesis that species richness is different from zero at the equator), or because the parameters are all interdependent (e.g. the fact that the `habitatforest` parameter is not significantly different from zero means nothing because it is balanced against the estimate for `habitatforest:latitude`). The one slightly meaningful inference is for the `habitatforest:latitude` parameter, which is telling us that there is little evidence for an interaction on the scale of the linear predictor. But even this doesn't necessarily translate to the biological hypothesis of no interaction between habitat and latitude since the relationship with latitude is nonlinear, as we saw in previous scripts. Take away: most of the hypothesis tests in these summary tables are usually irrelevant to hypotheses that we are scientifically interested in, except in infrequent special cases. If we want to test specific scientific hypotheses, we'll need to be deliberate about expressing them as alternative models.

95% confidence intervals for parameters are obtained by profiling the likelihood surface. This is just as we did earlier by varying one parameter while optimizing over the others. Here, the intervals are frequentist and are formed by assuming that the likelihood ratio has a $\chi ^{2}$ sampling distribution. Note this is not the same as the pure likelihood intervals we constructed earlier since here the likelihood ratio is merely a statistic and the inferential basis is the sampling distribution of the likelihood ratio.

```{r}
confint(fitHxL)
```

The correlation matrix for the parameters shows how the `habitatforest` and `habitatforest:latitude` parameters are highly interdependent.

```{r}
cov2cor(vcov(fitHxL))
```

Although meaningless for a single model, the log likelihood is useful for comparison with alternative models. Here's how we can extract the log likelihood.

```{r}
logLik(fitHxL)
```

To construct intervals for the curves, we can use the `predict()` function. For GLMs, there is no `interval="confidence"` option, so we have to construct intervals from the standard errors. This is approximate. More accurate intervals are obtained by parametric bootstrap.

```{r}
newd <- data.frame(latitude = rep(seq(41.92, 45, length.out=100), 2),
                   habitat = factor(rep(c("bog","forest"), each=100)))
preds <- predict(fitHxL, newdata=newd, se.fit=TRUE)
mnlp <- preds$fit        #mean of the linear predictor
selp <- preds$se.fit     #se of the linear predictor
cillp <- mnlp - 2 * selp #lower of 95% CI for linear predictor
ciulp <- mnlp + 2 * selp #upper
cil <- exp(cillp)        #lower of 95% CI for response scale
ciu <- exp(ciulp)        #upper
mu <- exp(mnlp)          #mean of response scale
preds <- cbind(newd,preds,cil,ciu,mu)
```

Plotting model with the data is as for the Bayesian GLM in the previous script.

```{r}
bfc <- c("#d95f02", "#1b9e77") #bog & forest colors
preds |>
    ggplot() +
    geom_ribbon(aes(x=latitude, ymin=cil, ymax=ciu, fill=habitat), alpha=0.2) +
    geom_line(aes(x=latitude, y=mu, col=habitat)) +
    geom_point(data=ant, aes(x=latitude, y=richness, col=habitat)) +
    annotate("text", x=42.7, y=3.3, label="Bog", col=bfc[1]) +
    annotate("text", x=43.85, y=9.5, label="Forest", col=bfc[2]) +
    scale_fill_manual(values=bfc) +
    scale_color_manual(values=bfc) +
    scale_y_continuous(breaks=seq(0, 20, 4), minor_breaks=seq(0, 20, 2)) +
    coord_cartesian(ylim=c(0, 18)) +
    xlab("Latitude (degrees north)") +
    ylab("Ant species richness") +
    theme(legend.position="none")
```



