---
title: Ant data Generalized Linear Mixed Model
author: Brett Melbourne
date: 11 Nov 2020 (minor updates 19 Nov 2024)
output: github_document
---
Seventh in a series of scripts to analyze the ant data described in Ellison (2004). This script fits multilevel models to fully account for the design structure. I'll demonstrate both frequentist and Bayesian approaches.

Set up (mostly for Bayesian analysis):

```{r results=FALSE, message=FALSE, warning=FALSE}
library(lme4)
library(ggplot2)
library(rstanarm)
library(bayesplot)
options(mc.cores=parallel::detectCores())
theme_set(theme_bw())
source("source/hpdi.R") #For calculating credible intervals
```

Read in and plot the data:

```{r}
ant <- read.csv("data/ants.csv")
ant$habitat <- factor(ant$habitat)
ant |>
    ggplot(mapping=aes(x=latitude, y=richness, col=habitat)) +
    geom_point() +
    ylim(0,18)
```

## GLMM
Recall that our first analysis was a GLM with Poisson distribution and log link. The multilevel model will do the same. In the design, at each site there are a pair of plots, one plot is a bog while the other is a forest.

![ants_sketch](images/ants_sketch.svg)

Thus, *site* is a grouping variable, while *plot* is the data scale. *Habitat type* is a predictor at the plot scale. *Latitude* is a predictor at the site scale.

```{r}
ant$site <- factor(ant$site)
```

We'll first try a maximum likelihood fit using `glmer` but we'll see that the algorithm fails to converge (although the failed fit is pretty close):

```{r}
glmerHxL <- glmer(richness ~ habitat + latitude + habitat:latitude + (1|site),
                 family=poisson, data=ant)
```

We get a warning suggesting to rescale variables. If we look at the correlation matrix (`Correlation of Fixed Effects`), we see a very high correlation (-1.000) between the `intercept` and `latitude` parameters, and the `habitat` and `habitat:latitude` parameters:

```{r}
summary(glmerHxL)
```

So, indeed this correlation might be the problem. Scaling and centering latitude fixes convergence.

```{r}
ant$latitude_s <- scale(ant$latitude)
glmerHxL <- glmer(richness ~ habitat + latitude_s + habitat:latitude_s + (1|site),
                 family=poisson, data=ant)
```

In the summary, we see we have a random effect term: the variance at the site scale, which is quite small.

```{r}
summary(glmerHxL)
```

Let's compare to the fit without the site grouping structure. We see that parameter estimates are almost the same.

```{r}
summary(glm(richness ~ habitat + latitude_s + habitat:latitude_s,
    family=poisson, data=ant))
```

So, why should we include the grouping term? First, because it is part of the sample design. Thus, we know it is a contributing factor. This is already a fact of the design and we can't explain it away, for example, with a non-significant hypothesis test for an effect of site. Second, we see that the autocorrelation due to site has an effect on the uncertainty of the latitude parameter. The mixed-effects model estimates the latitude effect (see `latitude_s` parameter) as smaller with more uncertainty (greater standard error), and a larger p-value for its null-hypothesis test. This is because there is effectively less replication of latitude because latitude occurs at the site scale and plots within sites are pseudoreplicates with respect to latitude. However, the correlation of plots within sites is not high (i.e. the random effect variance is low), so plots within sites seem mostly independent and hence these adjustments for dependence are relatively small.

Now turning to the Bayesian model. We have the same likelihood but we now have weakly informative priors as well (using here the sensible default priors). If we did this analysis with the scaled latitude variable we would see that the Bayesian analysis is substantially similar to the frequentist one. However, recall that in `rstanarm`, the predictor variables are rescaled automatically, so we'll use the unscaled latitude variable for convenience as it will return results on the original scale, saving us from carrying out the back-conversions. We see that the site random effect is estimated as almost the same as the likelihood estimate above.

```{r cache=TRUE}
bayesHxL <- stan_glmer(richness ~ habitat + latitude + habitat:latitude + (1|site), 
                       family=poisson, data=ant)
print(summary(bayesHxL)[,c("mean","sd","n_eff","Rhat")], digits=3)
```

Inspect diagnostics of the fit

```{r eval=FALSE}
launch_shinystan(bayesHxL)
```

In particular, the posterior distributions for the linear coefficients $\beta_i$ are all nicely symmetric, as expected for this type of model.

```{r}
mcmc_hist(as.array(bayesHxL), pars="(Intercept)", 
          regex_pars=c("latitude", "habitat"), bins=75)
```

We could next form credible intervals and prediction intervals but we'll defer that to the overdispersed model below.

## Overdispersed model

In the Poisson model, the variance is by definition equal to the mean. What if there is more variance than this? This situation is called overdispersion. We can include any extra variance by using a plot-level error term, which effectively makes the likelihood a Poisson-lognormal model. Our multilevel model now has two extra levels of variance, a plot-level variance and a site-level variance, in addition to the Poisson variance. Including a plot-level variance in a multilevel model is a very neat way to include extra variation in single parameter distributions like the Poisson. I prefer it compared to other options (such as using a negative binomial distribution) because the Normal error variance is more interpretable and can be compared directly to variances at other levels in the multilevel model.

First define a plot-level indicator. Plot is the lowest level in this dataset, so this is effectively the row number of the dataset.

```{r}
ant$plot <- factor(1:nrow(ant))
#ant$plot <- 1:nrow(ant) #works just as well not as factor
```

Fit the mixed effects model now including the plot random effect. The overdispersed GLMM using `glmer` with the unstandardized latitude again fails to converge.

```{r}
glmer_odHxL <- glmer(richness ~ habitat + latitude + habitat:latitude + (1|site) + (1|plot),
                        family=poisson, data=ant)
```

Failure to fit often happens with `glmer`. Given the error message, we might guess the trouble is with the high parameter correlations. So, let's try rescaling latitude:

```{r}
ant$latitude_s <- scale(ant$latitude)
glmer_odHxL_s <- glmer(richness ~ habitat + latitude_s + habitat:latitude_s + 
                       (1|site) + (1|plot), family=poisson, data=ant)
```

This fits successfully. There is a warning because the plot-scale variance estimate (the overdispersion term) is near zero (i.e. near the boundary of parameter space).

```{r}
summary(glmer_odHxL_s)
```
We could now go on to use this model for inference as before. For intervals, we'll need to use a parametric bootstrap approach. Instead, we'll now turn to the Bayesian model.


## Overdispersed model (Bayesian fit)

We don't have any such problems with the Bayesian model, which takes the overdispersion term in stride (the prior regularizes the estimate). I'm also going to up the iterations here to smooth out the credible intervals.

```{r cache=TRUE}
bayes_odHxL <- stan_glmer(richness ~ habitat + latitude + habitat:latitude + 
                          (1|site) + (1|plot), family=poisson, data=ant,
                          warmup=1000, iter=16000)
print(summary(bayes_odHxL)[,c("mean","sd","n_eff","Rhat")], digits=3)
```

Inspecting the posterior distribution for the parameters reveals nice symmetric distributions, except for the Sigma parameters (variances of the random effects), which is expected.

```{r eval=FALSE}
launch_shinystan(bayes_odHxL)
```

Sigma_plot in particular has a high probability mass near zero.

```{r}
mcmc_hist(as.array(bayes_odHxL), regex_pars=c("Sigma"), bins=75)
```


## Bayesian intervals

The code for the GLMM is substantially the same as the GLM that we looked at previously. First form a new dataset for prediction.

```{r}
newd <- data.frame(latitude=rep(seq(from=41.92, to=45, length.out=50), 2),
                   habitat=factor(rep(c("bog","forest"), each=50)))
```

Then derive samples for the posterior distribution of the inverse link function, i.e. Dist($\mu$), which we'll call `pmu`. In the GLMM we can choose whether we want predictions to include the specific grouping terms or not (in other words, do we want to predict for a specific site?). Here, we want to predict for generic new sites, so we don't include the specific sites. To do that, we use the argument `re.form=NA`, which means "don't form the random effects").

```{r message=FALSE}
pmu <- posterior_linpred(bayes_odHxL, transform=TRUE, re.form=NA, newdata=newd)
```

This is a matrix with samples in rows and the variable combinations in columns. The estimated means are then:

```{r}
mnmu <- colMeans(pmu)
```

and the 95% credible intervals for the mean are:

```{r}
n <- nrow(newd)
mean_intervals <- data.frame(mulo95=rep(NA,n), muhi95=rep(NA,n))
for ( i in 1:n ) {
    mean_intervals[i,] <- hpdi(pmu[,i], prob=0.95)
}
```

For predictions, first derive samples for the posterior predictive distribution, which we'll call ppd:

```{r}
ppd <- posterior_predict(bayes_odHxL, re.form=NA, newdata=newd)
```

and the prediction intervals (here CPI) are then:

```{r}
prediction_intervals <- predictive_interval(ppd, prob=0.95)
prediction_intervals <- data.frame(prediction_intervals)
names(prediction_intervals) <- c("ppdlo95", "ppdhi95")
```

The plot shows that the credible intervals for the means are a little wider than the fit that did not include the site-level grouping term or the overdispersion term (compare to [11_6_ants_bayes_GLM_rstanarm.md](11_6_ants_bayes_GLM_rstanarm.md)).

```{r}
preds <- cbind(newd, mnmu, mean_intervals, prediction_intervals)
bfc <- c("#d95f02", "#1b9e77") #bog & forest colors
preds |>
    ggplot() +
    geom_ribbon(aes(x=latitude, ymin=mulo95, ymax=muhi95, fill=habitat),
                alpha=0.2) +
    geom_line(aes(x=latitude, y=mnmu, col=habitat)) +
    geom_line(aes(x=latitude, y=ppdlo95, col=habitat), lty=2) +
    geom_line(aes(x=latitude, y=ppdhi95, col=habitat), lty=2) +
    geom_point(data=ant, aes(x=latitude, y=richness, col=habitat)) +
    annotate("text", x=42.7, y=3.3, label="Bog", col=bfc[1]) +
    annotate("text", x=43.85, y=9.5, label="Forest", col=bfc[2]) +
    scale_fill_manual(values=bfc) +
    scale_color_manual(values=bfc) +
    scale_y_continuous(breaks=seq(0, 20, 4), minor_breaks=seq(0, 20, 2)) +
    coord_cartesian(ylim=c(0, 20)) +
    xlab("Latitude (degrees north)") +
    ylab("Ant species richness") +
    theme(legend.position="none")
```

Now we can derive the differences in species richness between forest and bog across latitude, as previously (code is the same):

```{r}
diff <- pmu[,51:100] - pmu[,1:50]
diff_mn <- colMeans(diff)
n <- ncol(diff)
diff_cpi <- data.frame(difflo95=rep(NA,n), diffhi95=rep(NA,n))
for ( i in 1:n ) {
    diff_cpi[i,] <- quantile(diff[,i], prob=c(0.025,0.975))
}
diff_df <- data.frame(cbind(diff_mn, diff_cpi, latitude=newd$latitude[1:50]))
```

The plot shows similar estimates compared to the fit that did not include the site-level grouping term or the overdispersion term (compare to [11_6_ants_bayes_GLM_rstanarm.md](11_6_ants_bayes_GLM_rstanarm.md)). However, the decline with latitude is a little less steep and the credible interval is a little wider after accounting for the spatial structure and overdispersion.

```{r}
diff_df |> 
    ggplot() +
    geom_ribbon(mapping=aes(x=latitude, ymin=difflo95, ymax=diffhi95), alpha=0.2) +
    geom_line(mapping=aes(x=latitude, y=diff_mn)) +
    coord_cartesian(ylim=c(0,8)) +
    xlab("Latitude (degrees north)") +
    ylab("Difference in species richness (forest - bog)")
```

